### 2018-04-20 1_intro.tex
Cleanup a lot. Fix titles.

Research of weak memory models firstly aims to \textit{formalise} develop the formal approach of understanding programs with respect to weak memory models which is systematic, sound and complete. The first (and so far the only) such a framework was presented in 2010~\cite{alglave2010shared}.
%
Secondly, researchers work on extracting the WMMs for hardware architectures from existing implementations of from their specifications, that are written in natural language and thus suffer from ambiguities and incompleteness. Over last decade the memory models have been extracted for most mainstream multiprocessor architectures, such as x86-TSO (\textit{Total Store Order}) model for x86 architecture formalised in 2009~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding}, for Alpha~\cite{?} and Sparc~\cite{?} memory models were defined in the specification. Moreover, most modern high-level programming languages rely on relaxed memory model. Thus, the memory model for Java that is based on the \textit{happens-before} principle~\cite{lamport1978time} was introduced in J2SE 5.0 in 2004~\cite{manson2005java}; the C++11 standard~\cite{?} has introduced the set of hardware-independent synchronisation fences and atomic operations, whenever the C++17 memory model is based on the relation \textit{strongly happens-before}.
%batty2011mathematizing
Thirdly, important research direction targets the problem of verifying (or at least finding bugs in) existing software systems with respect to WMMs. Perhaps, the most notable work in this field is the project for defining the Linux kernel memory model, which is being actively developing these days~\cite{kernel1}.
%Distributed databases <also need the wmm, see transaction consistency~\cite{bailis2013highly}>

@@

Current work aims to rework the proof-of-concept tool \Porthos by extending the input language, which currently represents the minimum subset of C, and revising the general architecture of the tool in order to enhance performance, reliability and maintainability.
%interprocedural analysis
We called the new tool \textbf{the \tool{mousquetaires} framework} as now it not only performs the portability analysis, but can serve as the basis for SMT-solver driven weak memory model analysis.

>>>>

Research of weak memory models firstly aims to \textit{formalise} develop the formal approach of understanding programs with respect to weak memory models which is systematic, sound and complete. The first (and so far the only) such a framework was presented in 2010~\cite{alglave2010shared}.
In addition to developing rather theoretical basis, researchers work on extracting the WMMs for hardware architectures from existing implementations of from their specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over last decade the memory models have been defined for most mainstream multiprocessor architectures, such as x86-TSO and Sparc-TSO (for \textit{Total Store Order}) model for x86 and Sparc architecture formalised in 2009~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, and others.
There are projects for validating hardware architectures wrt. a memory model, e.g.~\cite{lustig2014pipecheck,lustig2016coatcheck}.

Most modern high-level programming languages rely on relaxed memory model as well.
Thus, the memory model for Java that is based on the \textit{happens-before} principle~\cite{lamport1978time} was introduced in J2SE 5.0 in 2004~\cite{manson2005java}; the C++11 standard~\cite{?} has introduced the set of hardware-independent synchronisation fences and atomic operations, whenever the C++17 memory model~\cite{batty2011mathematizing} is based on the relation \textit{strongly happens-before}. Weak memory are being formalised for even more abstract software environments, the notable project in this area is the project on formalising the Linux kernel memory model, which is being actively developing these days~\cite{kernel1}.
Furthermore, there is a wide range of tools that perform program verification wrt memory models (see~\cite{alglave2013software}, \cite{Porthos17}).

@@

Current work aims to rework the proof-of-concept tool \Porthos by extending the input language, which currently represents the minimum subset of C, and revising the general architecture of the tool in order to enhance performance, reliability and maintainability.
%interprocedural analysis
We called the new tool the~\tool{mousquetaires}~framework as now it not only performs the portability analysis, but can serve as the basis for SMT-solver driven weak memory model analysis.

### 2018-04-20 2_wmm.tex
Cleanup a lot. Fix titles.

\chapter{Memory Model-Aware Program Analysis}
\label{ch:wmm}

@@

%"The semantics of a program is a set of executions" https://johnwickerson.github.io/papers/memalloy.pdf

The classical approach to model concurrent programs is to use the \textit{global time}, a single order of interleavings among all events happened in different threads. However these models are easy to understand, it may be impossible to process \textit{all} possible states, number of which is exponentially large. However, there exist some equivalence classes such that the result of execution different interleavings from single equivalence class is the same (for instance, computations performed by a processor locally do not affect the global state).
%One way to do this is to use a non-deterministic computation-centric model~\cite{fri97}, such as ...%The idea in this class of models is based on the fact that the behaviour of a concurrent system is defined only by the interleavings of shared-memory operations, while being independent from the order of local computation events.
One such model is the \textit{event-based} representation of a program, which models a program as a directed graph (the \textit{event-flow graph}). The vertices of such a graph represent \textit{events} (simple low-level instructions; see Section~\ref{ch:wmm:model:events}), and edges represent \textit{relations} over the events (see Section~\ref{ch:wmm:model:relations}). Thus, a candidate execution constitutes the set of guesses which event (from the set of possible next events) is to be executed.

@@

The synchronisation instructions (fences) cause the \textit{barrier events}, that do not perform any computation or memory value transfer, instead, they add new relations to the program model that restrict the set of allowed behaviours. Semantically, a fence may serve as a synchronisation barrier, or as a flush of local memory caches, etc.

@@

%he read-from relation rf records the correspondance between the reads, the matching writes, and the communicated values on the computation Ξ //From cat.pdf
%The \textit{control-flow} instructions (conditional and unconditional jumps) are encoded into the model directly, without additional events, as the $po$-relation (for \textit{program order}; see Chapter~\ref{ch:wmm:model:relations} for detailed definition of relations).
\begin{itemize}
	\item
	The \textit{control-flow} of a program is defined by the \textit{program-order} relation \po~$\subset~\mathbb{E}~\times~\mathbb{E}$, which represents the total order of events of same process.
	For instance, if the instruction $i_1$ generates the event $e_1$ and the instruction $i_2$ follows $i_1$ and generates the event $e_2$, then $e_1 \xrightarrow{\mathtt{po}} e_2$. 
	%Some new relations may be acquired : dp, po-loc

	\item
	The \textit{data-flow} of a program is defined by \textit{communication relations}: the \textit{read-from} relation \rf~$\subset~\mathbb{W}~\times~\mathbb{R}$ that maps each write event to the read event that reads its value, the \textit{coherence order}-relation \co~$\subset~\mathbb{W}~\times~\mathbb{W}$
	%(also called \texttt{ws}-relation for \textit{write serialisation})
	defines the total order on writes to the same location across all processes.

	\item
	Events from the same process are related by the \textit{scope relation} \sr~$\subset~\mathbb{E}~\times~\mathbb{E}$. 
	% TODO: note that we do not use the scope tree, just labels
\end{itemize}

%The \textit{from-read order} \fr $\subset~\mathbb{R}~\times~\mathbb{W}$ that maps a read to possible writes preceding the current write event (this relation is the inversion of the \rf-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).
%TODO: perephrase last sentence! alglave thesis, p. 36

@@

%The semantics of a concurrent program is represented as the set of allowed executions.
An \textit{execution} (trace, run) of a program is an ordered set of events defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}. A \textit{candidate execution} is an execution that is not constrained by the memory model (i.e., the execution for that the validity is not checked). As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program at a whole.
%The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%the path in the event-flow graph. 
%An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}.

%An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$~\cite{alglave2010shared}.
The Figure~\ref{simple_wmm_x86_pic} illustrates four possible candidate executions for the litmus test Example~\ref{simple_wmm_x86} (the pictures are generated by the \texttt{herd7} tool, version 7.47). Since there are no conditional jumps, the \po-relation is defined and we do not need to guess it. Since each thread performs single write followed by a single read, the \co-relation is also defined (it relates the initial write event with the write event to the same location). Thus, there are only four possible executions defined by the choice of \rf-relation. The candidate executions pictured in Figures~\ref{simple_wmm_x86_pic:sub1}--\ref{simple_wmm_x86_pic:sub3} are consistent both under strong memory model SC and under relaxed memory models x86-TSO, Power, ARM, and others. However, the execution shown in Figure~\ref{simple_wmm_x86_pic:sub3} is still consistent under relaxed-memory architectures, but it becomes inconsistent under SC architecture as it forbids cycles over \fr$\cup$\po.
%The event $b$:\texttt{(Ry=0)} reads value \texttt{0} at the shared location \texttt{y} from the initial write event $e$\texttt{(Wy=0)} (the red edge of \texttt{rf}-relation), consequently, 
%the event $d$:\texttt{(Rx=0)} reads value \texttt{0} at the shared location \texttt{x} from the initial write event $f$\texttt{(Wx=0)} (the red edge of \texttt{rf}-relation). 
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

%\includegraphics[width=0.4\textwidth]{img/my/simple_wmm_x86.png}
\begin{figure}
\centering
\begin{subfigure}[t]{.28\textwidth}
  \centering
  \includegraphics[width=1.2\linewidth]{img/my/sb-example/SB-dot.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub1}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-2.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub2}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-3.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub3}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-4.png}
  \caption{Final state: \texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub4}
\end{subfigure}
\hfill
\caption{Candidate executions for the litmus test Example~\ref{simple_wmm_x86}}
\label{simple_wmm_x86_pic}
\end{figure}

\subsection{Weak Memory Models}
\label{ch:wmm:wmm}

\section{The \cat language}

Weak memory models can be described in the \cat language~\cite{alglave2016syntax}. It is the domain specific language for describing consistency properties of concurrent programs. The \cat language combines the expressive power of a functional language (it is inspired by OCaml and adopts its types, first-class functions, pattern matching and other features) with types, expressions and assertions specific for operating with relations and sets of executions.
% succinctly by constraining an abstraction of parallel program executions into a candidate execution and possibly extending this candidate execution with additional constraints on the execution environment.

%represents the functional language extended by the theory of  allows to axiomatically define derived relations, architecture-specific fences, and constraints over relations.

The definition of new relations  derived relations can be produced using the following operations over relations~\cite{alglave2016syntax}: % 3.1.6 Relations between events

\begin{itemize}
	\item \textit{the union} of two relations \texttt{r1} and \texttt{r2} is \texttt{r1\,|\,r2}
	\item \textit{the intersection} of two relations \texttt{r1} and \texttt{r2} is \texttt{r1\,\&\,r2}
	\item \textit{the difference} of two relations \texttt{r1} and \texttt{r2} is $\mathtt{r1\,\backslash\,r2}$
	\item \textit{the sequence}\footnote{The sequence of two relations \texttt{r1} and \texttt{r2} is defined as the set of pairs \texttt{(x,y)} such that there exists an intervening \texttt{z}, such that $\mathtt{(x,z)} \in \mathtt{r1}$ and $\mathtt{(z,y)} \in \mathtt{r2}$}
	of two relations \texttt{r1} and \texttt{r2} is \texttt{r1;r2}
\end{itemize}

@@

%There is a barrier instruction \texttt{mfence} that may be used for flushing the buffers into the main memory.

briefly some hw memory models: X86-TSO, Alpha, POWER, ;

language memory models: Java, C++;

library-level kernel memory model, ref to github with tests

@@

\section{Some examples of WMM}

//axioms of TSO wmm

// example of sets: rf, co, ... for the code snipped used before

>>>>

\chapter{Memory model-aware analysis}
%\chapter{A generic framework}
\label{ch:wmm}

@@

%"The semantics of a program is a set of executions" https://johnwickerson.github.io/papers/memalloy.pdf
The classical approach for modeling concurrent programs is to use the \textit{global time}, a single order of interleavings among all events happened in different threads.
Although these models are easy to understand, it may be impossible to process \textit{all} possible states, number of which is exponentially large.
However, there exist equivalence classes such that the result of execution different interleavings from single equivalence class is the same (for instance, computations performed by a processor locally do not affect the global state).
%One way to do this is to use a non-deterministic computation-centric model~\cite{fri97}, such as ...%The idea in this class of models is based on the fact that the behaviour of a concurrent system is defined only by the interleavings of shared-memory operations, while being independent from the order of local computation events.
One such model is the \textit{event-based} representation of a program, which models a program as a directed graph of events (the \textit{event-flow graph}).
The vertices of such a graph represent \textit{events} (independent low-level instructions; see Section~\ref{ch:wmm:model:events}), and edges represent \textit{relations} over the events (see Section~\ref{ch:wmm:model:relations}).

@@

The synchronisation instructions (fences) cause the \textit{barrier events}, that do not perform any computation or memory value transfer, instead, they add new relations to the program model that restrict the set of allowed behaviours.
Functionally, a fence may be a synchronisation barrier or a instruction of flushing the local memory caches, etc.

@@

\vspace{1em}
The basic relations are the following~\cite{alglave2010shared}:

\begin{itemize}
    \item The \textit{control-flow} of a program is defined by the \textit{program-order} relation \po~$\subset~\mathbb{E}~\times~\mathbb{E}$, which represents the total order of events of same process.
	For instance, if the instruction $i_1$ generates the event $e_1$ and the instruction $i_2$ follows $i_1$ and generates the event $e_2$, then $e_1 \relation{po} e_2$.
	%Some new relations may be acquired : dp, po-loc

    \item The \textit{data-flow} of a program is defined by \textit{communication relations}:
        \begin{itemize}[noitemsep]
            \item the \textit{read-from} relation \rf~$\subset~\mathbb{W}~\times~\mathbb{R}$ that maps each write event to the read event that reads its value;
            \item the \textit{coherence order} relation \co~$\subset~\mathbb{W}~\times~\mathbb{W}$ that defines the total order on writes to the same location across all processes         (also called the \textit{write serialisation} \texttt{ws}-relation);
        \end{itemize}

	\item Events from the same process are related by the \textit{scope relation} \sr~$\subset~\mathbb{E}~\times~\mathbb{E}$.
    In contrast to the \tool{herd} tool, the \mousquetaires does not use hierarchy of scopes (depicted as the scope tree); instead, it uses simple lables that indicate which process has produced certain event.
\end{itemize}

%\vspace{0.5em}
Below we enumerate some derived relations~\cite{alglave2010shared}:

\begin{itemize}

    \item the \textit{from-read} relation $\fr \subset~\mathbb{R}~\times~\mathbb{W}$ that maps a read to writes preceding the write event from which the read reads the value: \\
    $r \relation{fr} w = (\exists w' . w' \relation{rf} r \land w' \relation{co} w)$.

    \item the \textit{data dependency} relation \texttt{dp}, which is a subset of \po-relation that always has a read at its source (it connects the read to the write which it depends on).
    %$\mathtt{com} = (\mathtt{rf} \cup \mathtt{co} \cup \mathtt{fr})$

    \item the \textit{external} (and \textit{internal}) \textit{read-from} relations that restrict the \rf-relation to the different (respectively, same) processes.

    \item the \texttt{po-loc} relation that is the \po-relation over events that access to the same shared variable:
    $m_1 \relation{po-loc} m_2 = (m_1 \relation{po} m_2 \land \mathtt{loc}(m_1) = \mathtt{loc}(m_2))$.

    \item the semantics of fences and barriers specific for different architectures may be defined as derived relations.

    %"Preserved program order defines the set of reorderings guaranteed by the architecture not to occur, based on the types of the accesses." http://delivery.acm.org/10.1145/2750000/2742219/p635-lustig.pdf

\end{itemize}

The work~\cite{alglave2010shared} (Chapter~2) provides definition of basic properties of relations, such as \textit{reflexivity} and \textit{irreflexivity}, \textit{transitivity} and \textit{transitive closure}, \textit{acyclicity}.
Thereafter, weak memory models make asserts over these properties, thus restricting the set of allowed behaviours of the system.

% TODO: some derived relations
%The \textit{from-read order} \fr $\subset~\mathbb{R}~\times~\mathbb{W}$ that maps a read to possible writes preceding the current write event (this relation is the inversion of the \rf-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).
%TODO: perephrase last sentence! alglave thesis, p. 36

@@

The \textit{candidate execution} is a path in the event-flow graph defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}. %the sequence of choices from guesses which event is to be executed next.
%The semantics of a concurrent program is represented as the set of allowed executions.
%An \textit{execution} (trace, run) of a program is an ordered set of events defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}.
%A \textit{candidate execution} is an execution that is not constrained by the memory model (i.e., the execution for that the validity is not checked).
%As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program at a whole.
%The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%the path in the event-flow graph.
%An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}.
%An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$~\cite{alglave2010shared}.
Figure~\ref{simple_wmm_x86_pic} illustrates four possible candidate executions for the litmus test Example~\ref{simple_wmm_x86} (the pictures are generated by the \tool{herd7} tool, version~7.47).
Since there are no conditional jumps, the \po-relation is defined and we do not need to guess it.
Since each thread performs single write followed by a single read, the \co-relation is also defined (it relates the initial write event with the write event to the same location).
Thus, there are only four possible executions defined by the choice of \rf-relation.
The candidate executions pictured in Figures~\ref{simple_wmm_x86_pic:sub1}--\ref{simple_wmm_x86_pic:sub3} are consistent both under strong memory model SC and under relaxed memory models x86-TSO, Power, ARM, and some others.
However, the execution shown in Figure~\ref{simple_wmm_x86_pic:sub3} is still consistent under relaxed-memory architectures, but it becomes inconsistent under SC architecture as it forbids cycles over~$\fr\cup\po$.
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

\begin{figure}[H]
\centering
\begin{subfigure}[t]{.28\textwidth}
  \centering
  \includegraphics[width=1.2\linewidth]{img/my/sb-example/SB-dot.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub1}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-2.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub2}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-3.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub3}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-4.png}
  \caption{Final state: \texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub4}
\end{subfigure}
\hfill
\caption{Possible candidate executions for the litmus test Example~\ref{simple_wmm_x86}}
\label{simple_wmm_x86_pic}
\end{figure}

\section{The \cat language}
\label{ch:wmm:cat}

Weak memory models are defined via the \cat language~\cite{alglave2016syntax}.
This is a domain specific language for describing consistency properties of concurrent programs.
The \cat language combines expressive power of a functional language (it is inspired by OCaml and adopts its types, first-class functions, pattern matching and other features) with types, expressions and assertions that are specific for operating with relations and executions.
%represents the functional language extended by the theory of  allows to axiomatically define derived relations, architecture-specific fences, and constraints over relations.

The derived relations can be defined via the keyword \texttt{let} and the following operations over relations~\cite{alglave2016syntax}: % 3.1.6 Relations between events

\begin{itemize}
	\item \textit{the union} of two relations \relname{r1} and \relname{r2} is \relname{r1\,|\,r2}
	\item \textit{the intersection} of two relations \relname{r1} and \relname{r2} is \relname{r1\,\&\,r2}
	\item \textit{the difference} of two relations \relname{r1} and \relname{r2} is $\mathtt{r1\,\backslash\,r2}$
	\item \textit{the sequence}\footnote{The sequence of two relations \relname{r1} and \relname{r2} is defined as the set of pairs \relname{(x,y)} such that there exists an intervening \relname{z}, such that $\relname{(x,z)} \in \relname{r1}$ and $\relname{(z,y)} \in \relname{r2}$}
	of two relations \relname{r1} and \relname{r2} is \relname{r1;r2}
\end{itemize}

For instance, Figure~\ref{example:x86-cat} contains part of x86-TSO model~\cite{owens2009better} that asserts acyclicity of communication relation and \relname{po-loc} relation:

\begin{figure}[!h]
\begin{lstlisting}
...
let com = rf | fr | co
let po-loc = po & loc
acyclic po-loc | com
\end{lstlisting}
\caption{Excerpt from the x86-TSO memory model in \cat language}
\label{example:x86-cat}
\end{figure}

@@

%There is a barrier instruction \relation{mfence} that may be used for flushing the buffers into the main memory.

%briefly some hw memory models: X86-TSO, Alpha, POWER, ;

%language memory models: Java, C++;

%library-level kernel memory model, ref to github with tests

@@

%\section{Some examples of WMM}

%//axioms of TSO wmm

%// example of sets: rf, co, ... for the code snipped used before
